Symbolic Abstract Data Type Inference

### 2015-11-04 draft.tex
Addressing Constantin’s comments.

\authorinfo{Constantin Enea}
           {Université Paris Diderot, France}
           {cenea@univ-paris-diderot.fr}

@@

% \acks …

>>>>

\authorinfo{Constantin Enea}
           {Université Paris Diderot, France}
           {cenea@liafa.univ-paris-diderot.fr}

@@

% \acks …
% \section*{Acknowledgements}

\acks This work is supported in part by the VECOLIB project (ANR-14-CE28-0018),
the N-GREENS Software project (Ref. S2013/ICE-2731), and AMAROUT-II
(EU-FP7-COFUND-291803).

### 2015-11-04 histories.tex
Addressing Constantin’s comments.

    \item read $\Rightarrow v$ operations are read-only, and match themselves
    when they return $v = -$, or a write$(v)$ operation, if one
    exists, and otherwise have no match,

>>>>

    \item read $\Rightarrow v$ operations are read-only, and match themselves
    when they return $v = -$ (empty), or a write$(v)$ operation, if one
    exists, and otherwise have no match,

### 2015-11-04 inference.tex
Addressing Constantin’s comments.

A Set $K$ \emph{generates} $H$ when the closure of $K$ under the relation
$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$
is equal to $H$, i.e., $H = \set{ h : \exists h' \in K.\ h' \to^\ast h }$.
A \emph{kernel} of a set $H$ is a minimal set generating $H$. While the kernels
of arbitrary sets need not be unique, the kernels of sets which
have sequential kernels are unique. Furthermore, Section~\ref{sec:nature}
demonstrates that the histories of naturally-occurring implementations have
unique kernels, which we assume for the remainder of this work.
An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
of histories of some implementation $\mathcal{I}$; we say that \emph{$A$ is the
ADT of $\mathcal{I}$}.

>>>>

A set $K$ \emph{generates} $H$ when the closure $K^*$ of $K$ under the relation
$\mathord\to = (\to_\mathrm{o} \cup \to_\mathrm{p} \cup \to_\mathrm{c})$
is equal to $H$, i.e., $H = \set{ h : \exists h' \in K.\ h' \to^\ast h }$.
A \emph{kernel} of a set $H$ is a minimal set generating $H$. While the kernels
of arbitrary sets need not be unique, the kernels of sets which
have sequential kernels are unique. Furthermore, Section~\ref{sec:nature}
demonstrates that the histories of naturally-occurring implementations have
unique kernels, which we assume for the remainder of this work.
An \emph{abstract data type (ADT)} $A$ is the kernel of the set $H(\mathcal{I})$
of histories of some implementation $\mathcal{I}$; we say that \emph{$A$ is the
ADT of $\mathcal{I}$}.

### 2015-11-04 intro.tex
Addressing Constantin’s comments.

Although formal ADT specifications are indispensable for scalable program
reasoning, formal-specification writing is a burden for which few programmers
posses the required combination of expertise and willingness to overcome.
Typically programmers write simple ADT \emph{reference implementations},
e.g.,~whose methods are synchronized via a global lock, and refine them with
more efficient fine-grained implementations, e.g.,~reducing synchronization
bottlenecks using specialized hardware instructions such as atomic compare and
swap (CAS).

>>>>

Although formal ADT specifications are indispensable for scalable program
reasoning, formal-specification writing is a burden for which few programmers
possess the required combination of expertise and willingness to overcome.
Typically programmers write simple ADT \emph{reference implementations},
e.g.,~whose methods are synchronized via a global lock, and refine them with
more efficient fine-grained implementations, e.g.,~reducing synchronization
bottlenecks using specialized hardware instructions such as atomic compare and
swap (CAS).

### 2015-11-04 nature.tex
Addressing Constantin’s comments.

Atomic queues and stacks implement atomic collections of data values with
first-in-first-out (FIFO) and last-in-first-out (LIFO) removal order,
respectively, providing two methods:\footnote{These methods are also known as
enqueue and dequeue, or push and pop.}
\begin{itemize}

@@

An execution \emph{inserts unique values} if the argument value to each insert
operation which returns $-$ is unique. We give a faithful matching
scheme over executions which insert unique values as follows:
\begin{itemize}

@@

\end{itemize}
Unlike atomic queues, the work-stealing queue permits values to be removed
twice: once normally, via the take operation, and once exceptionally, via the
steal operation. A faithful matching scheme over executions which add
unique values is analogous to the scheme for stacks and queues: give
operations match themselves, while take and steal operations match the give
operation which added their returned value, or themselves, in case the nil
value is returned. Note that the work-stealing queue ADT has width $2$,
since a pairs of concurrent take and steal operations returning the same value
are permitted, while sequentially they are not. As our implementation is
currently limited to width-$1$ histories (see the discussion in
Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.

We compute the following twenty-four histories to generate the complement of
the work-stealing queue ADT:
\begin{verbatim}
  [1:_] take => 1  #             [1:_] steal => 1  #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)          #
  [2:2] take => - (RO)    #      [2:2] steal => - (RO)    #
  ---                            ---
  [1:2] take => 1  #             [1:2] steal => 1  #
  [2:2] give(1)      #           [2:2] give(1)       #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] steal => 2      #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:1] take => 1    #           [2:1] take => 1     #
  [3:1] take => 1      #         [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)         #
  [2:2] take => - (RO)    #      [2:2] take => - (RO)    #
  [3:1] take => 1           #    [3:1] steal => 1          #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:1] steal => 1    #          [2:1] steal => 1    #
  [3:1] take => 1       #        [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)          #       [1:1] give(1)          #
  [2:2] steal => - (RO)    #     [2:2] steal => - (RO)    #
  [3:1] take => 1            #   [3:1] steal => 1           #
  ---                            ---
  [1:2] take => 1   #            [1:3] take => 1   #
  [2:2] give(1)       #          [2:3] steal => 1    #
  [3:2] steal => 1      #        [3:3] give(1)         #
  ---                            ---
  [1:2] steal => 1  #            [1:3] steal => 1  #
  [2:2] give(1)       #          [2:3] take => 1     #
  [3:2] take => 1       #        [3:3] give(1)         #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] take => 2       #
  [4:1] take => 1        #       [4:1] steal => 1        #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:2] give(2)       #          [2:2] give(2)       #
  [3:2] steal => 2      #        [3:2] steal => 2      #
  [4:1] take => 1         #      [4:1] steal => 1        #
\end{verbatim}
As new generators are discovered for $n=1$ and $n=2$ matches only, the
work-stealing queue ADT is predictable.

>>>>

Atomic queues and stacks implement atomic collections of data values with
first-in-first-out (FIFO) and last-in-first-out (LIFO) removal order,
respectively, providing two methods:
\begin{itemize}

@@

An execution \emph{inserts unique values} if the argument value to each insert
operation returning $-$ is unique. We give a faithful matching
scheme over executions which insert unique values as follows:
\begin{itemize}

@@

\end{itemize}
Unlike atomic queues, the work-stealing queue permits values to be removed
twice: once normally, via the take operation, and once exceptionally, via the
steal operation. A faithful matching scheme over executions which add
unique values is analogous to the scheme for stacks and queues: give
operations match themselves, while take and steal operations match the give
operation which added their returned value, or themselves, in case the nil
value is returned. Note that the work-stealing queue ADT has width $2$,
since pairs of concurrent take and steal operations returning the same value
are permitted, while sequentially they are not. As our implementation is
currently limited to width-$1$ histories (see the discussion in
Section~\ref{sec:discussion}), below we generate only the width-$1$ complement.

We compute the following twenty-four histories to generate the complement of
the work-stealing queue ADT:
\begin{verbatim}
  [1:X] take => 1  #             [1:X] steal => 1  #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)          #
  [2:2] take => - (RO)    #      [2:2] steal => - (RO)    #
  ---                            ---
  [1:2] take => 1  #             [1:2] steal => 1  #
  [2:2] give(1)      #           [2:2] give(1)       #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] steal => 2      #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:1] take => 1    #           [2:1] take => 1     #
  [3:1] take => 1      #         [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)         #        [1:1] give(1)         #
  [2:2] take => - (RO)    #      [2:2] take => - (RO)    #
  [3:1] take => 1           #    [3:1] steal => 1          #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:1] steal => 1    #          [2:1] steal => 1    #
  [3:1] take => 1       #        [3:1] steal => 1      #
  ---                            ---
  [1:1] give(1)          #       [1:1] give(1)          #
  [2:2] steal => - (RO)    #     [2:2] steal => - (RO)    #
  [3:1] take => 1            #   [3:1] steal => 1           #
  ---                            ---
  [1:2] take => 1   #            [1:3] take => 1   #
  [2:2] give(1)       #          [2:3] steal => 1    #
  [3:2] steal => 1      #        [3:3] give(1)         #
  ---                            ---
  [1:2] steal => 1  #            [1:3] steal => 1  #
  [2:2] give(1)       #          [2:3] take => 1     #
  [3:2] take => 1       #        [3:3] give(1)         #
  ---                            ---
  [1:1] give(1)    #             [1:1] give(1)     #
  [2:2] give(2)      #           [2:2] give(2)       #
  [3:2] take => 2      #         [3:2] take => 2       #
  [4:1] take => 1        #       [4:1] steal => 1        #
  ---                            ---
  [1:1] give(1)     #            [1:1] give(1)     #
  [2:2] give(2)       #          [2:2] give(2)       #
  [3:2] steal => 2      #        [3:2] steal => 2      #
  [4:1] take => 1         #      [4:1] steal => 1        #
\end{verbatim}
As new generators are discovered for $n=1$ and $n=2$ matches only, the
work-stealing queue ADT is predictable.

### 2015-11-04 patterns.tex
Addressing Constantin’s comments.

%!TEX root = draft.tex
\section{Finite ADT Representations}
\label{sec:patterns}

In this section we demonstrate that naturally-occurring ADTs can be precisely
represented by finite sets of histories, despite the fact that these ADTs admit
infinite sets of histories. This results relies on the identification of certain
algebraic properties of the sets of histories admitted by ADTs. In particular,
we find that sets of histories admitted by ADTs are closed under the removal
of certain operations, and that these sets adhere to a well-founded ordering
under the relation induced by such removals.

@@

  The following four histories generate the complement of the atomic
  single-value register ADT, witnessing either an unmatched read operation:
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  a read of an uninitialized register occurring after some write:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] read => - (RO)    #
\end{verbatim}
  a read which happens before its matching write operation:
\begin{verbatim}
  [1:2] read => 1 (RO)  #
  [2:2] write(1)          #
\end{verbatim}
  and a read matching a write which is not the most recent:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] write(2)          #
  [3:1] read => 1 (RO)      #
\end{verbatim}
  Every sequential history not admitted by the atomic register ADT embeds at
  least one of these four histories.

>>>>

\section{Finite ADT Representations}
\label{sec:patterns}

In this section we demonstrate that naturally-occurring ADTs can be precisely
represented by finite sets of histories, despite the fact that these ADTs admit
infinite sets of histories. This result relies on the identification of certain
algebraic properties of the sets of histories admitted by ADTs. In particular,
we find that sets of histories admitted by ADTs are closed under the removal
of certain operations, and that these sets adhere to a well-founded ordering
under the relation induced by such removals.

@@

  The following four histories generate the complement of the atomic
  single-value register ADT, witnessing either an unmatched read operation:
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  a read of an uninitialized register occurring after some write:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] read => - (RO)    #
\end{verbatim}
  a read which happens before its matching write operation:
\begin{verbatim}
  [1:2] read => 1 (RO)  #
  [2:2] write(1)          #
\end{verbatim}
  and a read matching a write which is not the most recent:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] write(2)          #
  [3:1] read => 1 (RO)      #
\end{verbatim}
  Every sequential history not admitted by the atomic register ADT embeds
  (via $\succeq$) at least one of these four histories.
